---
title: "Precision Oncology Mutation Enrichment Analysis"
subtitle: "Response-associated somatic mutation screening with low-frequency and mutation-burden safeguards"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(20260716)
```

## Study question

This analysis asks whether nonsynonymous somatic mutations are enriched in drug responders versus non-responders, using MAF files and sample-level clinical response annotations.

This is an exploratory precision-oncology analysis. Associations reported here should be interpreted as candidate hypotheses unless validated in an independent cohort or supported by orthogonal clinical and biological evidence.

## Assumptions

- Input data are available under `vanallen-assessment/`.
- MAF files are stored in `vanallen-assessment/mafs/`.
- Clinical annotations are stored in `vanallen-assessment/sample-information.tsv`.
- The clinical file contains `Tumor_Sample_Barcode`, `Response`, and `Nonsynonymous_mutations_per_Mb`.
- `Response` contains two response groups, expected to be `Responder` and `Non-Responder`.
- Each `Tumor_Sample_Barcode` is treated as one analyzable tumor sample.
- The repository does not include the assessment data; only the analysis workflow and archived rendered output are versioned.

## Packages

```{r packages}
required_packages <- c(
  "dplyr",
  "forcats",
  "ggplot2",
  "here",
  "knitr",
  "purrr",
  "readr",
  "stringr",
  "tibble",
  "tidyr"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Install missing packages before knitting: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(dplyr)
library(forcats)
library(ggplot2)
library(here)
library(knitr)
library(purrr)
library(readr)
library(stringr)
library(tibble)
library(tidyr)
```

## Load and harmonize input data

```{r load-data}
base_path <- here("vanallen-assessment")
mafs_path <- file.path(base_path, "mafs")
clinical_path <- file.path(base_path, "sample-information.tsv")

if (!dir.exists(mafs_path)) {
  stop("MAF directory not found: ", mafs_path)
}

if (!file.exists(clinical_path)) {
  stop("Clinical file not found: ", clinical_path)
}

maf_files <- list.files(mafs_path, pattern = "\\.maf(\\.gz)?$", full.names = TRUE)

if (length(maf_files) == 0) {
  stop("No .maf or .maf.gz files found under: ", mafs_path)
}

raw_maf <- maf_files |>
  set_names(basename(maf_files)) |>
  map_dfr(
    ~ read_tsv(.x, comment = "#", show_col_types = FALSE),
    .id = "source_maf"
  )

clinical <- read_tsv(clinical_path, show_col_types = FALSE)

required_maf_cols <- c("Hugo_Symbol", "Tumor_Sample_Barcode", "Variant_Classification")
required_clinical_cols <- c(
  "Tumor_Sample_Barcode",
  "Response",
  "Nonsynonymous_mutations_per_Mb"
)

missing_maf_cols <- setdiff(required_maf_cols, colnames(raw_maf))
missing_clinical_cols <- setdiff(required_clinical_cols, colnames(clinical))

if (length(missing_maf_cols) > 0) {
  stop("MAF is missing required columns: ", paste(missing_maf_cols, collapse = ", "))
}

if (length(missing_clinical_cols) > 0) {
  stop("Clinical file is missing required columns: ", paste(missing_clinical_cols, collapse = ", "))
}

clinical <- clinical |>
  mutate(
    Response = factor(Response),
    Nonsynonymous_mutations_per_Mb = as.numeric(Nonsynonymous_mutations_per_Mb)
  )

sample_summary <- clinical |>
  count(Response, name = "n_samples")

kable(sample_summary, caption = "Samples by response group")
```

## Filter to nonsynonymous variants

Silent variants are excluded before gene-level enrichment testing. A gene is considered mutated in a sample if at least one nonsynonymous variant is observed in that gene for that sample.

```{r filter-variants}
nonsyn_maf <- raw_maf |>
  filter(Variant_Classification != "Silent") |>
  semi_join(clinical, by = "Tumor_Sample_Barcode")

gene_sample <- nonsyn_maf |>
  distinct(Tumor_Sample_Barcode, Hugo_Symbol)

analysis_samples <- clinical |>
  semi_join(
    tibble(Tumor_Sample_Barcode = unique(raw_maf$Tumor_Sample_Barcode)),
    by = "Tumor_Sample_Barcode"
  )

n_samples <- n_distinct(analysis_samples$Tumor_Sample_Barcode)
n_genes <- n_distinct(nonsyn_maf$Hugo_Symbol)
n_variants <- nrow(nonsyn_maf)

tibble(
  analyzable_samples = n_samples,
  nonsynonymous_variants = n_variants,
  genes_with_nonsynonymous_variants = n_genes
) |>
  kable(caption = "Analysis set after nonsynonymous filtering")
```

## Most frequently mutated genes

Large genes are often recurrently mutated because of size and mutability rather than disease-specific selection. The FLAGS list is used only as a cautionary label, not as an automatic exclusion rule.

```{r top-genes}
flags <- c(
  "TTN", "MUC16", "OBSCN", "AHNAK2", "SYNE1", "FLG", "MUC5B",
  "DNAH17", "PLEC", "DST", "SYNE2", "NEB", "HSPG2", "LAMA5",
  "AHNAK", "HMCN1", "USH2A", "DNAH11", "MACF1", "MUC17"
)

gene_frequency <- gene_sample |>
  count(Hugo_Symbol, name = "mutated_samples") |>
  mutate(
    mutation_prevalence = mutated_samples / n_samples,
    flag_gene = Hugo_Symbol %in% flags
  ) |>
  arrange(desc(mutated_samples), Hugo_Symbol)

top_15 <- gene_frequency |>
  slice_head(n = 15)

kable(top_15, digits = 3, caption = "Top 15 genes by mutated sample count")

flagged_top_15 <- intersect(top_15$Hugo_Symbol, flags)

tibble(
  flagged_genes_in_top_15 = ifelse(
    length(flagged_top_15) == 0,
    "None",
    paste(flagged_top_15, collapse = ", ")
  )
) |>
  kable(caption = "Large-gene caution labels among the top 15")
```

## Gene-level enrichment by response

For each gene, a 2 x 2 Fisher exact test compares mutation presence versus response group. This is appropriate for small sample sizes and sparse mutation counts.

Low-frequency mutations are reported but separated from the primary screen because singletons and doubletons can create unstable odds ratios.

Operational tiers:

- `primary_screen`: gene mutated in at least 5 samples.
- `low_frequency`: gene mutated in 2-4 samples.
- `singleton`: gene mutated in 1 sample.

The threshold is intentionally conservative for a 50-sample cohort. It does not imply that lower-frequency events are unimportant; it prevents sparse nominal findings from being promoted as biomarker candidates without additional evidence.

```{r enrichment}
response_levels <- levels(droplevels(analysis_samples$Response))

if (length(response_levels) != 2) {
  stop(
    "Expected exactly two response groups for Fisher exact testing. Observed: ",
    paste(response_levels, collapse = ", ")
  )
}

preferred_responder_label <- "Responder"
if (!preferred_responder_label %in% response_levels) {
  preferred_responder_label <- response_levels[2]
}

test_gene <- function(gene_symbol) {
  mutated_samples <- gene_sample |>
    filter(Hugo_Symbol == gene_symbol) |>
    pull(Tumor_Sample_Barcode)

  test_df <- analysis_samples |>
    mutate(
      mutated = factor(
        Tumor_Sample_Barcode %in% mutated_samples,
        levels = c(FALSE, TRUE)
      ),
      Response = factor(Response, levels = response_levels)
    )

  contingency <- table(test_df$mutated, test_df$Response)
  fisher_result <- fisher.test(contingency)

  counts <- test_df |>
    count(Response, mutated) |>
    complete(Response, mutated = c(FALSE, TRUE), fill = list(n = 0)) |>
    unite("response_mutated", Response, mutated, sep = "_") |>
    pivot_wider(names_from = response_mutated, values_from = n)

  tibble(
    Hugo_Symbol = gene_symbol,
    mutated_samples = length(mutated_samples),
    odds_ratio = unname(fisher_result$estimate),
    p_value = fisher_result$p.value,
    responder_label = preferred_responder_label
  ) |>
    bind_cols(counts)
}

enrichment_results <- gene_frequency |>
  pull(Hugo_Symbol) |>
  map_dfr(test_gene) |>
  mutate(
    q_value = p.adjust(p_value, method = "BH"),
    frequency_tier = case_when(
      mutated_samples >= 5 ~ "primary_screen",
      mutated_samples >= 2 ~ "low_frequency",
      TRUE ~ "singleton"
    )
  ) |>
  arrange(p_value, desc(mutated_samples), Hugo_Symbol)

primary_screen <- enrichment_results |>
  filter(frequency_tier == "primary_screen")

low_frequency_findings <- enrichment_results |>
  filter(frequency_tier != "primary_screen", p_value < 0.05)

kable(
  primary_screen |> slice_head(n = 20),
  digits = 4,
  caption = "Primary enrichment screen: genes mutated in at least five samples"
)

kable(
  low_frequency_findings |> slice_head(n = 20),
  digits = 4,
  caption = "Low-frequency nominal findings: shown for review, not treated as definitive"
)
```

```{r enrichment-plot, fig.width=8, fig.height=6}
plot_df <- enrichment_results |>
  filter(mutated_samples >= 2) |>
  mutate(
    neg_log10_p = -log10(p_value),
    Hugo_Symbol = fct_reorder(Hugo_Symbol, neg_log10_p)
  ) |>
  slice_min(p_value, n = 30)

ggplot(
  plot_df,
  aes(
    x = mutated_samples,
    y = neg_log10_p,
    color = frequency_tier,
    label = Hugo_Symbol
  )
) +
  geom_point(size = 3, alpha = 0.85) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  scale_x_continuous(trans = "log2") +
  labs(
    title = "Mutation-response enrichment screen",
    subtitle = "Fisher exact tests; x-axis log2-scaled for readability",
    x = "Mutated samples",
    y = "-log10 nominal p-value",
    color = "Frequency tier"
  ) +
  theme_minimal(base_size = 12)
```

## Inspect the top candidate

The leading candidate is selected from the primary screen by nominal p-value. If no primary-screen gene is available, the report stops rather than over-interpreting low-frequency findings.

```{r top-candidate}
if (nrow(primary_screen) == 0) {
  stop("No genes met the primary-screen frequency threshold.")
}

top_candidate <- primary_screen |>
  slice_min(p_value, n = 1, with_ties = FALSE)

top_gene <- top_candidate$Hugo_Symbol[[1]]

kable(top_candidate, digits = 4, caption = "Top response-associated candidate")
```

```{r top-candidate-burden}
top_mutant_samples <- gene_sample |>
  filter(Hugo_Symbol == top_gene) |>
  pull(Tumor_Sample_Barcode)

burden_df <- analysis_samples |>
  mutate(
    mutation_status = if_else(
      Tumor_Sample_Barcode %in% top_mutant_samples,
      paste0(top_gene, " mutant"),
      paste0(top_gene, " wild-type")
    ),
    mutation_status = factor(
      mutation_status,
      levels = c(paste0(top_gene, " mutant"), paste0(top_gene, " wild-type"))
    )
  ) |>
  filter(!is.na(Nonsynonymous_mutations_per_Mb))

burden_summary <- burden_df |>
  group_by(mutation_status) |>
  summarise(
    n = n(),
    median_tmb = median(Nonsynonymous_mutations_per_Mb),
    mean_tmb = mean(Nonsynonymous_mutations_per_Mb),
    iqr_tmb = IQR(Nonsynonymous_mutations_per_Mb),
    .groups = "drop"
  )

burden_test <- wilcox.test(
  Nonsynonymous_mutations_per_Mb ~ mutation_status,
  data = burden_df,
  exact = FALSE
)

kable(burden_summary, digits = 3, caption = "Mutation burden by top candidate status")

tibble(
  test = "Wilcoxon rank-sum",
  p_value = burden_test$p.value
) |>
  kable(digits = 4, caption = "Mutation burden comparison")
```

```{r burden-plot, fig.width=7, fig.height=5}
ggplot(
  burden_df,
  aes(x = mutation_status, y = Nonsynonymous_mutations_per_Mb, fill = mutation_status)
) +
  geom_boxplot(alpha = 0.55, outlier.shape = NA) +
  geom_jitter(width = 0.12, alpha = 0.75, size = 2) +
  labs(
    title = paste("Mutation burden by", top_gene, "mutation status"),
    x = NULL,
    y = "Nonsynonymous mutations per Mb"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## Sensitivity: response association adjusted for mutation burden

Mutation burden can confound response-associated gene findings, especially for DNA-repair genes or hypermutated samples. A small-sample logistic model is included as a sensitivity analysis, not as a definitive predictive model.

```{r burden-adjusted-sensitivity}
sensitivity_df <- analysis_samples |>
  mutate(
    response_binary = Response == preferred_responder_label,
    top_gene_mutated = Tumor_Sample_Barcode %in% top_mutant_samples
  ) |>
  filter(!is.na(Nonsynonymous_mutations_per_Mb))

glm_fit <- glm(
  response_binary ~ top_gene_mutated + Nonsynonymous_mutations_per_Mb,
  data = sensitivity_df,
  family = binomial()
)

glm_summary <- summary(glm_fit)$coefficients |>
  as.data.frame() |>
  rownames_to_column("term") |>
  as_tibble() |>
  rename(
    estimate = Estimate,
    std_error = `Std. Error`,
    z_value = `z value`,
    p_value = `Pr(>|z|)`
  )

kable(
  glm_summary,
  digits = 4,
  caption = "Sensitivity model: response association adjusted for mutation burden"
)
```

## Interpretation

The original knitted output reported `ERCC2` as the most significantly enriched gene, with 9 mutant samples and 41 wild-type samples. It also reported higher mean nonsynonymous mutations per Mb in the `ERCC2`-mutant group than in the wild-type group.

This re-analysis keeps `ERCC2` as a candidate only if it remains significant under the primary frequency threshold and multiple-testing review. The key expert interpretation is cautious:

- `ERCC2` is biologically plausible in treatment-response contexts because it participates in nucleotide excision repair.
- Higher mutation burden among `ERCC2`-mutant samples may represent a biological mechanism, a confounder, or both.
- Low-frequency mutations should be surfaced for review but not promoted as validated biomarkers.
- Any responder-enriched gene from this cohort should be treated as hypothesis-generating until externally validated.
